---
title: 'Reproducible Research: Peer Assessment 2 - Major Weather Events Analysis'
author: "Dimitris Zabelis"
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: true
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3         
    number_sections: true
    highlight: tango
    latex_engine: xelatex
---
```{r, echo=FALSE, results='asis'}
if (knitr::is_latex_output()) cat("\\newpage")
```

# Synopsis



```{r, echo=FALSE, results='asis'}
if (knitr::is_latex_output()) cat("\\newpage")
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(data.table)
```

# Data Acquisition

The "repdata_data_StormData.csv" file holds 902,297 rows and 37 columns. Hence, data.table::fread() will be used to read in the data, which is faster than read.csv() and read_csv().

```{r}
dt <- fread(cmd = "bzip2 -dc repdata_data_StormData.csv.bz2", 
            header = T, sep = ",", showProgress = F)
```

# Data Profiling

The inspection will be held through the prism of the relevant columns for this project:

- BGN_DATE, which represents the day the weather event started.
- EVTYPE, which represents the type of the weather event.
- FATALITIES & INJURIES, which are the counts of direct human deaths and injuries attributed to the event, respectively.
- PROPDMG & PROPDMGEXP, which are the estimated property damage and its exponent, respectively.
- CROPDMG & CROPDMGEXP, which are the estimated crop damage and its exponent, respectively.

<u>Important Note</u>:

- As explained in the [Storm Data FAQ Page](https://www.ncei.noaa.gov/stormevents/faq.jsp)(Section:"How are tornadoes counted?"), the health and economic consequences for **multi-county events** (major weather events crossing many counties) are segmented. In other words, when an event crosses many counties the **consequences** of that event are **spread across the respective rows of the different counties the event crosses**. That way, all multi-county events are treated as **separate events** that occur in different counties, where each row that belongs to the same multi-county event carries its own consequences (instead of recycled aggregates representing the total amounts). Hence, during the preprocessing stage, grouping and aggregation will occur so that the overall impact of each event type can be calculated.

```{r}
relevant_column_names <- c("BGN_DATE","EVTYPE", "FATALITIES", "INJURIES", 
                           "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
```

Transposed summary of the data.table:
```{r}
glimpse(dt[, ..relevant_column_names])
```

NAs:
```{r}
naniar::miss_var_summary(dt[, ..relevant_column_names])
```
The 8 relevant columns do not contain any NAs.

Check for the existence of duplicate records:
```{r}
janitor::get_dupes(dt)
```
There are no duplicates in "dt".

Data summary by relevant variable:
```{r}
skimr::skim(dt[, ..relevant_column_names])
```

Observations from the summary statistics:

- The "EVTYPE" column holds 985 unique values, whereas the number of major weather events is significantly  smaller (48). This points to naming inconsistencies.
- Most values in the "PROPDMGEXP" and "CROPDMGEXP" columns are empty, signifying that most of the provided numbers in "PROPDMG" and "CROPDMG" are multiplied by 1 (dollars, rather than thousands or millions of dollars)
- All four numeric columns follow a zero-inflated distribution, that is, a distribution that has a substantial spike at zero and a long right tail of non-zero values.

Unique values of the "PROPDMGEXP" column:
```{r}
unique(dt$PROPDMGEXP)
```

Unique values of the "CROPDMGEXP" column:
```{r}
unique(dt$CROPDMGEXP)
```

# Data Preprocessing

## Description of the Preprocessing Pipeline

Steps in the Preprocessing Pipeline:

1. Rename the Columns
2. Preprocess the "BGN_DATE" column
   - Convert to Date
   - Filter out records before 01/01/1996
3. Clean the "EVTYPE" Column
   - Remove trailing and leading whitespaces
   - Convert to title case
   - Use str_extract() for automatic and bulk transformation of most unique 
   values of the column.
   - Use fcase() to manually group the unique values from the output of 
   str_extract().
   - Filter out unique values associated with data entry errors or general terms. 
4. Create two derived columns holding the total property and agricultural damages 
   - Multiply the damages by their multiplier, which is calculated by 
   applying the "exponent_to_multiplier" function to the "PROPDMGEXP" and 
   "CROPDMGEXP" columns.
5. Remove the NAs step 4. produced.
6. Group by the "EVTYPE" column and calculate summary statistics.

## Note regarding the "BGN_DATE" column

Why filter out records before 01/01/1996?

- Because [NOAA](https://www.ncei.noaa.gov/stormevents/faq.jsp) didn't begin recording all weather event types until that date.
- More specifically, before 1996, only Tornado, Thunderstorm Wind, and Hail data were collected.
- The recording of all other weather events presented in the [storm data event table](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)(p.6) began after 1996.
- Considering that, including dates before 1996 could potentially skew the analysis toward these event types.

## Supporting Objects

### Related to str_extract()

The values of both the "types" and "extra_types" vectors that follow will be used as patterns for the extraction and transformation of all 985 unique values of the "EVTYPE" column through the utilization of the stringr::str_extract() function.

The "types" vector holds all 48 unique weather event types from the [storm data event table](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)(p.6):
```{r}
types <- c(
  "Astronomical Low Tide", "Avalanche", "Blizzard", "Coastal Flood",
  "Cold/Wind Chill", "Debris Flow", "Dense Fog", "Dense Smoke",
  "Drought", "Dust Devil", "Dust Storm", "Excessive Heat",
  "Extreme Cold/Wind Chill", "Flash Flood", "Frost/Freeze",
  "Funnel Cloud", "Freezing Fog","Heavy Rain", "Heavy Snow", "High Surf",
  "Hurricane \\(Typhoon\\)", "Ice Storm", "Lake-Effect Snow", "Lakeshore Flood",
  "Lightning", "Marine Hail", "Marine High Wind", "Marine Strong Wind",
  "Marine Thunderstorm Wind", "Rip Current", "Seiche", "Sleet",
  "Storm Surge/Tide", "Strong Wind", "Thunderstorm Wind", "Tornado",
  "Tropical Depression", "Tropical Storm", "Tsunami", "Volcanic Ash", "High Wind",
  "Waterspout", "Wildfire", "Winter Storm", "Winter Weather", "Flood", "Hail", "Heat"
)
```

The "extra_types" vector contains the patterns required to extract the remaining unique values of the "EVTYPE" column:
```{r}
extra_types <- c(
        #--- Rare Entries -----------------------------------------------------#
        "Dam", "Typhoon", "Volcanic", "Drowning","Smoke", "Vog", "Turbulence",
        "Hypothermia", "Hyperthermia", "Remnants", "Northern Lights", "Landslump", 
        "Rock Slide", "Excessive", "Driest", "None", "Wnd", "Apache", "Southeast", 
        "Metro", "Red Flag", "Wall Cloud", "Freezing Drizzle", "Freezing Spray",
        #--- Misspellings -----------------------------------------------------#
        "Lighting", "Ligntning", "Tunderstorm", "Thundestorm", "Thuderstorm",
        "Thundeerstorm", "Thunerstorm", "Mircoburst", "Micoburst", "Floo", 
        "Avalance", "Torndao",
        #--- Single-Word Patterns ---------------------------------------------#
        "Summary", "\\?", "Other", "Tstm", "Snow", "Rain",
        "Cold", "Hurricane", "Thunderstorm", "Ice", "Freeze", 
        "Frost", "Fire", "Mud", "Funnel", "Fog", "Microburst", "Warm", "Dry",
        "Wet", "Hot", "Cool", "Tide", "Surf", "Swell", "Wave", "Surge", "Dust", 
        "Spout", "Precip","Glaze", "Mix", "Urban", "Erosion", "Record", "Temp", 
        "Landslide", "Shower", "Marine", "Downburst", "Gustnado", "Mild", "Sea", 
        "Stream", "Beach", "Water", "Icy", "Storm", "Wind", "Severe", "High"
)
```

Regex Matching Notes:

- "Spout" will match both "Waterspout" and "Landspout".
- "Precip" will match "Precipitation" and the misspelled "Precipatation".
- "Mix" will match "Wintry Mix", "Wintery Mix", "Winter Mix", and "Heavy Mix".
- "Marine" will match "Marine Mishap" and "Marine Accident".
- "Sea" will match "High Seas", "Heavy Seas", and "Rough Seas".
- "Stream" will match "Small Stream" and "Sml Stream Fld".
- "Beach" will match "Beach Erosion" and the misspelled "Beach Erosin".
- "Water" will match "High Water" and "Rapidly Rising Water".
- "Icy" will match "Icy Roads" ("Ice" ≠ "Icy").
- "Red Flag", will match "Red Flag Fire Wx" and "Red Flag Criteria".

Pattern for regex matching:
```{r}
pattern <- str_c(c(types, extra_types), collapse = "|")
```

### Related to the Recoding of the "EVTYPE" Column

The following vectors are used to group similar weather events (with fcase()) to further reduce the number of unique values of the "EVTYPE" column after the application of the stringr::str_extract() function:
```{r}

thunderstorm_vals <- c(
  "Thunderstorm Wind", "Thunderstorm", "Tstm", "Marine Thunderstorm Wind",
  "Tunderstorm", "Thundestorm", "Thuderstorm", "Thundeerstorm", "Thunerstorm",
  "Microburst", "Mircoburst", "Micoburst", "Downburst"
)

lightning_vals <- c("Lightning","Lighting", "Ligntning")

tornado_vals <- c(
  "Tornado", "Waterspout", "Funnel Cloud", "Funnel",
  "Spout", "Wall Cloud", "Gustnado", "Torndao", "spout"
)

flood_vals <- c(
  "Flash Flood", "Flood", "Lakeshore Flood", "Coastal Flood", "flood",
  "Storm Surge/Tide", "Surge", "Dam", "Urban", "Stream", "Water", "Floo"
)

wind_vals <- c("High Wind", "Strong Wind", "Wind", "Wnd", "wind", 
               "Marine High Wind", "Marine Strong Wind")

hail_vals <- c("Marine Hail", "Hail")

winter_snow_vals <- c(
  "Heavy Snow", "Blizzard", "Winter Storm", "Winter Weather",
  "Lake-Effect Snow", "Snow", "snow" ,"Ice Storm", "Sleet",
  "Ice", "Glaze", "Mix", "Icy"
)

cold_vals <- c(
  "Cold/Wind Chill", "Extreme Cold/Wind Chill", "Cold", "Cool",
  "Hypothermia", "Frost/Freeze", "Frost", "Freeze"
)

hurricane_tropical_vals <- c(
  "Hurricane (Typhoon)", "Typhoon", "Hurricane",
  "Tropical Storm", "Tropical Depression", "Remnants"
)

heat_vals <- c("Excessive Heat", "Heat", "Hot", "Warm", "Hyperthermia")

drought_vals <- c("Drought", "Dry", "Driest")

fog_vals <- c("Dense Fog", "Fog", "Freezing Fog")

smoke_vals <- c("Dense Smoke", "Smoke")

rain_vals <- c("Heavy Rain", "Rain", "Shower", "Wet", 
               "Precip", "Freezing Drizzle")

wildfire_vals <- c("Wildfire", "Fire")

avalanche_vals <- c("Avalanche", "Avalance")

landslide_vals <- c("Landslide", "Debris Flow", "Mud", "Rock Slide", "Landslump")

dust_vals <- c("Dust Devil", "Dust Storm", "Dust")

volcanic_vals <- c("Volcanic Ash", "Volcanic","Vog")

marine_coastal_vals <- c(
  "Rip Current", "High Surf", "Surf", "Swell", "Wave", "Astronomical Low Tide",
  "Sea", "sea", "Beach", "Tide", "Freezing Spray", "Tsunami", "Seiche",
  "Marine", # "Marine Mishap" and "Marine Accident"
  "Erosion" # Coastal Erosion
)
```

Note that NOAA separates "Thunderstorm Wind" and "Lightning" on the basis that they
are different hazards, even though both categories belong to the same storm system. Regarding the former, the health or economic consequences are caused by the wind of the storm. Regarding the latter, the health or economic consequences are caused by a lightning strike.

Note that marinal weather events are spread across different groups. This should be changed for marine vs land analysis, but suffices for this project.

- Most marinal weather events are part of the "Marine/Coastal" group. 
- "Marine Thunderstorm Wind" is part of the "Thunderstorm" group.
- "Marine High Wind" and "Marine Strong Wind" are part of the "wind" group. 
- "Marine Hail" is part of the "Hail" group.

The following strings are values that are not grouped (that is, values that pass through fcase() unchanged via the default parameter):

- "Drowning", "Northern Lights", "Storm", "Severe","High", "Mild", "Record"
- "Temp", "None", "Summary", "Other", "Apache", "Southeast", "Metro", "Excessive"
- "Red Flag", "Turbulence"

These values mostly represent data entry errors or very generic terms that could be attached to many different weather events.

All of the above values that do not belong to a group are filtered out except for the "Turbulence" value.

The "filter_out_vals" vector holds the abovementioned values (excluding "Turbulence"):
```{r}
filter_out_vals <- c(
        "Drowning", "Northern Lights", "Storm", "Severe","High", "Mild", "Record", 
        "Temp", "None", "Summary","Other", "Apache", "Southeast", "Metro", 
        "Excessive", "Red Flag", "storm"
) 
```

Recoding Function for the "EVTYPE" column:
```{r}
recode_event_type <- function(event_type){
        fcase(
                event_type %chin% thunderstorm_vals,       "Thunderstorm Wind",
                event_type %chin% lightning_vals,          "Lightning",
                event_type %chin% tornado_vals,            "Tornado",
                event_type %chin% flood_vals,              "Flood",
                event_type %chin% wind_vals,               "Wind",
                event_type %chin% hail_vals,               "Hail",
                event_type %chin% winter_snow_vals,        "Winter Storm",
                event_type %chin% cold_vals,               "Cold",
                event_type %chin% hurricane_tropical_vals, "Hurricane/Typhoon",
                event_type %chin% heat_vals,               "Heat",
                event_type %chin% drought_vals,            "Drought",
                event_type %chin% fog_vals,                "Fog",
                event_type %chin% smoke_vals,              "Smoke",
                event_type %chin% rain_vals,               "Heavy Rain",
                event_type %chin% wildfire_vals,           "Fire",
                event_type %chin% avalanche_vals,          "Avalanche",
                event_type %chin% landslide_vals,          "Landslide",
                event_type %chin% dust_vals,               "Dust Storm",
                event_type %chin% volcanic_vals,           "Volcanic Ash",
                event_type %chin% marine_coastal_vals,     "Marine/Coastal",
                default = event_type
        )
}
```

### Related to the Recoding of the "PROPDMGEXP" and "CROPDMGEXP" Columns

Recoding Function for the "PROPDMGEXP" and "CROPDMGEXP" columns:
```{r}
exponent_to_multiplier <- function(exp){
        fcase(
                exp %chin% c("H" , "h"), 1e2,
                exp %chin% c("K" , "k"), 1e3,
                exp %chin% c("M", "m"), 1e6,
                exp %chin% c("B"), 1e9,
                default = 0 # all digits from 0 to 8, "+" , "?", "-", & ""
        )
}
```

## Preprocessing Pipeline

```{r}
setnames(dt, relevant_column_names, 
         c("begin_date","event_type", "fatalities", "injuries", "property_damage", 
           "property_damage_exponent", "crop_damage", "crop_damage_exponent")
         )

invisible(dt[ 
        , `:=`(
                begin_date = as_date(mdy_hms(begin_date)),
                event_type = str_to_title(str_trim(event_type, side = "both"))
        )
])
```

```{r}
results <- dt[
        begin_date >= mdy("01/01/1996")
][
        , event_type := str_extract(event_type, regex(pattern, ignore_case = T))
][
        , event_type := recode_event_type(event_type)
][
        !(event_type %chin% filter_out_vals)
][
        , `:=`(
                total_property_damage = 
                        property_damage * 
                        exponent_to_multiplier(property_damage_exponent),
                total_crop_damage = 
                        crop_damage * 
                        exponent_to_multiplier(crop_damage_exponent)
        )
][
        , .(
                count = .N,
                fatalities_per_type = sum(fatalities),
                injuries_per_type = sum(injuries),
                total_property_damage_per_type = sum(total_property_damage),
                total_crop_damage_per_type = sum(total_crop_damage)
        ), by = event_type
][
        , `:=`(
                health_consequences_per_type = fatalities_per_type + injuries_per_type,
                damages_per_type = 
                        total_property_damage_per_type + total_crop_damage_per_type 
        )
][
        order(-health_consequences_per_type, -damages_per_type)
]
```

```{r}
unique(results$event_type)
```

```{r}
results
```


# Results

presentation of main results

# Software Environment

```{r}
sessionInfo()
```